Modelling state-switching functional data with hidden Markov models

### 2020-11-13 Report/Sections/models.tex
shorten intro

@@

Consider a sequence of $T$ curves (or dives, in our context). For any particular $t = 1,
\ldots,T$, dive $t$ is characterized by a possibly high-dimensional curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of $T^*_t$ within-curve (or fine-scale) observations, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$. The sequence $Y^*_{t}$ is made up of function values from curve $t$. We call the sequence of coarse-scale observations $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$ and the collection of all fine-scale observations $Y^* \equiv \big\{Y^*_1,\ldots,Y^*_T \big\}$.

Hidden Markov models describe state-switching Markovian processes in discrete time and will be the core structure of the models we use for $Y$ and $Y^*$. We first detail the structure of a traditional HMM. We then review a variation described by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations. Either the HMM or the CarHMM can be used to model a sequence of observations at one of the two scales: for the coarse scale sequence $Y$ or for each of the fine scale sequences, $Y_1^*,\ldots, Y_T^*$. To simultaneously model $Y$ and $Y^*$, we use the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}.
%which is comprised of two distinct,x but related, HMMs that jointly model these time series. 
Next, we describe how to reduce fine-scale correlation by transforming raw observations into meaningful summary statistics. In particular, we apply a discrete Fourier transform (DFT) to the components of $Y^*$ using a moving window to summarize its periodic behaviour. Finally, we show how each of these models can be altered and combined to form a wide variety of new models.

@@

An HMM is comprised of a sequence of unobserved states $X \equiv \big\{X_1, \ldots, X_T\big\}$ where $X_t$ is associated with the observation $Y_t$. Here we focus on the coarse scale sequence of observations $Y_1,\ldots, Y_T$, but an HMM can also be used as a sub-model of the fine-scale sequence for a particular dive. The $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time. 
The $X_t$'s form a Markov chain and take possible values $1, \ldots, N$. Their distribution is governed by the distribution of the initial state $X_1$ and the $N \times N$ transition probability matrix $\Gamma$, where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$. 
%
We assume that $X_1$ follows the chain's stationary distribution, which is denoted by $\delta \in \bbR^N$, with $i^{th}$ component
$\delta_i = \Pr(X_1 = i).$
A Markov chain's stationary distribution is determined by its probability transition matrix via $\delta = \delta \Gamma$ and $\sum_{i=1}^N \delta_i = 1$.
%
The distribution of an emission $Y_t$ depends only on the corresponding state $X_t$ and no other observations or hidden states: $p\left(y_t|\{X_1,\ldots, X_T\},\{Y_1,\ldots, Y_T\}/ \{Y_t\}\right) = p(y_t|X_t)$.
%
These conditional distributions are governed by state-dependent parameters. Namely, if $X_t = i$ then the state-dependent parameter is $\theta^{(i)}$ and we denote the conditional distribution of $Y_t$ given $X_t=i$ by its conditional density or probability mass function, denoted $f^{(i)}(\cdot ; \theta^{(i)})$, or sometimes simply $f^{(i)}(\cdot)$.
%
Figure \ref{fig:models}a represents the dependence structure of an HMM.

@@

In many applications where data are collected at high frequencies, intricate dependency structures may arise within the fine-scale process which cannot be adequately modelled with any of the HMM-based models mentioned thus far. For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour. To handle these additional fine-scale structures, we recommend replacing $Y_t^* = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant statistics that summarize this non-Markovian behaviour. To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over the elements $Y_t^*$. Subject-area experts are often required to determine the optimal window size and the specific summary statistics used. For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width $h$ and stride $h$ across $Y^*_t$. Namely:
%
\begin{align*}
    DFT\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right)
\end{align*}
%
for $t^* = 1,h+1,2h+1,\ldots$ and for $k = 0, 1, \ldots, h-1$. If $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise. Next, we index each window with $\tilde{t}^*$ and calculate the transformed observations $\Z_{t,\tilde{t}^*} \equiv \{\Zone_{t,\tilde{t}^*},\Ztwo_{t,\tilde{t}^*}\}$, where
%
\begin{equation}
    \label{eqn:z}
    \Zone_{t,\tilde{t}^*} \equiv \frac{1}{h}\sum_{n=1}^{h}Y^*_{t,h(\tilde{t}^*-1)+n}, \quad \Ztwo_{t,\tilde{t}^*} \equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,h(\tilde{t}^*-1)+1},\ldots, Y^*_{t,h\tilde{t}^*}\}(k)\Big|\Big|^2.
\end{equation}
%
In words, $\Zone_{t,\tilde{t}^*}$ is the average acceleration vector over window $\tilde{t}^*$ and $\Ztwo_{t,\tilde{t}^*}$ is the squared two-norm of the component of the window that can be attributed to frequencies between one and $\tilde{\omega} \leq h-1$ periods per window (by the Plancherel theorem). More intuitively, $\Ztwo_{t,\tilde{t}^*}$ corresponds to the ``wiggliness" of the acceleration data for dive $t$ within window $\tilde{t}^*$. The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with the help of a subject-matter expert. Other DFT-based choices for window summary statistics are possible, including the dominant frequency and amplitude. Setting the stride length equal to the window size $h$ avoids the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting. By summarizing the data using a relevant transformation, this method reduces correlation in the raw observations due to any complex dependence structure.

The fine-scale HMM must be redefined since $\Z_t$ exists on a coarser scale than $Y^*_t$ itself. In particular, the length of $\Z_t$ is denoted as $\tilde{T}_t^*$ and there are only $\tilde{T}_t^*$ hidden states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$, denoted as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$. The fine-scale probability transition matrices  $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ are then applied directly to $\tilde{X}^*_t$ and $\Z_t$ rather than $X^*_t$ and $Y^*_t$. It then follows that the likelihood of this model is identical to that of the original HMM but with $Y^*$ replaced with $\Z$ and $X^*$ replaced with $\tilde{X^*}$. In order to clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X}$ as hidden states as an HMM-DFT.

>>>>

% !TeX root = ../main.tex

@@

Consider a sequence of $T$ curves where for any particular $t = 1,
\ldots,T$, curve $t$ is characterized by a curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of $T^*_t$ within-curve (or fine-scale) observations $Y^*_{t}$. Namely, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up of function values from curve $t$. We call the sequence of coarse-scale observations $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$ and the collection of all fine-scale observations $Y^* \equiv \big\{Y^*_1,\ldots,Y^*_T \big\}$. We first detail the structure of a traditional HMM followed by three variations which generalize its base structure. We then show how each of these generalized models can be synthesized to form a wide variety of more complicated models.

%Hidden Markov models describe state-switching Markovian processes in discrete time and is the core structure we use to model $Y$ and $Y^*$. We first detail the structure of a traditional HMM, followed by the conditionally auto-regressive hidden Markov model (CarHMM) \citep{Lawler:2019}, a variation which explicitly models auto-correlation between observations. 
%Either the HMM or the CarHMM can model either the coarse scale sequence $Y$ or any of the fine scale sequences, $Y_1^*,\ldots, Y_T^*$. 
%We then describe a hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019} to simultaneously model $Y$ and $Y^*$. 
%, which is comprised of two distinct, but related, HMMs that jointly model these time series. 
%Next, we outline a method to reduce fine-scale correlation by transforming raw observations into meaningful summary statistics. In particular, we apply a discrete Fourier transform (DFT) to the components of $Y^*$ using a moving window to capture its periodic behaviour. Finally, we show how each of these models can be altered and combined to form a wide variety of new models.

@@

Hidden Markov models describe state-switching Markovian processes in discrete time and are the core structure we use to model both $Y$ and $Y^*$. For simplicity we focus on $Y$ to introduce the model. An HMM is comprised of a sequence of unobserved states $X \equiv \big\{X_1, \ldots, X_T\big\}$ together with an observation sequence $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$, where $X_t$ is associated with the observation $Y_t$. The $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time. 
The $X_t$'s form a Markov chain and can take values $1, \ldots, N$. Their distribution is governed by the distribution of the initial state $X_1$ and the $N \times N$ transition probability matrix $\Gamma$, where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$. 
%
We assume that $X_1$ follows the chain's stationary distribution, which is denoted by $\delta \in \bbR^N$ with $i^{th}$ component
$\delta_i = \Pr(X_1 = i).$
A Markov chain's stationary distribution is determined by its probability transition matrix via $\delta = \delta \Gamma$ and $\sum_{i=1}^N \delta_i = 1$.
%
The distribution of an emission $Y_t$ depends only on the corresponding state $X_t$ and no other observations or hidden states: $p\left(y_t|\{X_1,\ldots, X_T\},\{Y_1,\ldots, Y_T\}/ \{Y_t\}\right) = p(y_t|X_t)$.
%
These conditional distributions are governed by state-dependent parameters. Namely, if $X_t = i$ then the state-dependent parameter is $\theta^{(i)}$ and we denote the conditional distribution of $Y_t$ given $X_t=i$ by its conditional density or probability mass function, denoted $f^{(i)}(\cdot ; \theta^{(i)})$, or sometimes simply $f^{(i)}(\cdot)$.
%
Figure \ref{fig:models}a represents the dependence structure of an HMM.

@@

In many applications where data are collected at high frequencies, intricate dependency structures may arise within the fine-scale process which cannot be adequately modelled with any of the HMM-based models mentioned thus far. For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour. We recommend replacing $Y_t^* = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant summary statistics to describe any fine-scale non-Markovian behaviour. To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over the elements $Y_t^*$. Subject-area experts are often required to determine the optimal window size and the specific summary statistics used. For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width and stride length $h$ across $Y^*_t$:
%
\begin{align*}
    DFT\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right)
\end{align*}
%
for each $t^* = 1,h+1,2h+1,\ldots$ and $k = 0,1,\ldots, h-1$. If $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise. We then index each window with $\tilde t^*$ and calculate $\Z_{t,t^*} \equiv \{\Zone_{t,\tilde t^*},\Ztwo_{t,\tilde t^*}\}$, where
%
\begin{equation}
\begin{split}
    \label{eqn:z}
    \Zone_{t,\tilde t^*} &\equiv \frac{1}{h}\sum_{n=1}^{h}Y^*_{t,h(\tilde t^*-1)+n} \\
    %
    \Ztwo_{t,\tilde t^*} &\equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,h(\tilde t^*-1) + 1},\ldots, Y^*_{t,h\tilde t^*}\}(k)\Big|\Big|^2.
\end{split}
\end{equation}
%
The collection of $\Z_{t,\tilde t^*}$ for all windows is $\Z_t \equiv \{\Z_{t,1},\ldots,\Z_{t,\tilde T_t^*}\}$, where $\tilde T_t^* = \lfloor T^*_t/h \rfloor$. The hidden states corresponding to $\Z_t$ are distinct from those corresponding to $Y^*_t$ and are defined as $\tilde X^*_t \equiv \{\tilde X^*_{t,1},\ldots,\tilde X^*_{t,\tilde T_t^*}\}$.

In words, $\Zone_{t,\tilde t^*}$ is the average value over window $\tilde t^*$ and $\Ztwo_{t,\tilde t^*}$ is the squared two-norm of the component of window $\tilde t^*$ that can be attributed to frequencies between one and $\tilde{\omega} \leq h-1$ periods per window (by the Plancherel theorem). More intuitively, $\Ztwo_{t,\tilde t^*}$ corresponds to the ``wiggliness" of the acceleration data during dive $t$ and window $\tilde t^*$. The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with the help of a subject-matter expert. Other DFT-based choices for window summary statistics are possible, including the dominant frequency and amplitude. Transforming $Y^*_t$ to $\Z_t$ using a non-overlapping window transformation both avoids introducing artificial correlation and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting. By summarizing the data using a relevant transformation, the method overall also reduces correlation in the raw observations by summarizing complex dependence structures within each window. Figure \ref{fig:models}d represents the process of transforming $Y^*_{t}$ into $\Z_{t}$ within an HMM, including down-sampling.

%When down-sampling, we redefine the fine-scale probability transition matrices such that $\Gamma^{*(i)}_{jk} = \Pr(X^*_{t,t^*+h} = k | X^*_{t,t^*} = j, X_t = i)$ for all $t$ and $t^* = 1,h+1,2h+1,\ldots$. If the fine-scale model is a CarHMM, we also redefine the probability density function $f$ such that $f^{*(i,i^*)}(\cdot|\z_{t,t^*-100}; \theta^{*(i,i^*)})$ describes the density of $\Z_{t,t^*}$ given $X_t = i$, $X^*_{t,t^*} = i^*$, and $\Z_{t,t^*-100} = \z_{t,t^*-100}$.
%
%Once $\Z_t$ is calculated, it simply replaces the original fine-scale observations $Y^*_t$ in the fine-scale HMM structure. As such, the emission distributions and the 
The likelihood of an HMM with transformed fine-scale observations is identical to the original HMM but with $\Z_t$ and $\tilde X^*_t$ replacing $Y^*_t$ and $X^*_t$, respectively. In order to clearly differentiate models, we refer to an HMM with $\Z$ as observations as an HMM-DFT.

### 2020-11-13 Report/Sections/results.tex
shorten intro

We then select a model corresponding to the fine-scale observations of acceleration. Similarly to the coarse model, we rely on lag plots and visual inspection to select $N^*=3$. Although $N^*$ is selected heuristically, we test the validity of this model using techniques in Section \ref{subsec:model_validation}.
%
In contrast to the coarse-scale observations, the fine-scale acceleration data exhibit significant sinusoidal behaviour. Thus, we transform each fine-scale observation sequence $y_t^*$ into $\z_t$ using Equation (\ref{eqn:z}) with window size $h=100$ (two seconds) and maximum frequency $\tilde{\omega}=10$ (5 Hz). 
%We select a window size of two seconds since it appears be the smallest window size that captures at least one full period of each ``wiggle" in the acceleration data. 
We then have that $\z_{t,t^*} = \left\{\zone_{t,t^*},\ztwo_{t,t^*}\right\}$, where $\zone_{t,t^*}$ is a three dimensional vector of component-wise average acceleration and $\ztwo_{t,t^*}$ is a scalar describing the ``wiggliness" of a particular window. We also down-sample each observation $\z_t$ such that $\z_t = \left\{\z_{t,1},\z_{t,101},\z_{t,201},\ldots\right\}$ and $x^*_t = \left\{x^*_{t,1},x^*_{t,101},x^*_{t,201},\ldots\right\}$ to avoid overlapping windows. Even after transforming and down-sampling the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material). Therefore, we choose a CarHMM as defined in Section 2.2 as the fine-scale model.

@@

All components of $\Zone_{t,t^*}$ are assumed to be Normally distributed as in Equation (\ref{eqn:carhmm}) for all $t$ and $t^*$. Each component is assumed to have its own mean and variance parameters, but all components have the same auto-correlation parameter. 

@@

Given that $X^*_{t,t^*} = i^*$, $\Ztwo_{t,t^*}$ is assumed to follow a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$. We do not include $\Ztwo_{t,t^*-100}$ in the distribution of $\Ztwo_{t,t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state. 

@@

Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $Z_{t,t^*}$ when conditioned on $X^*_{t,t^*} = i^*$. 

@@

We refer to this final model as the \textbf{CarHHMM-DFT} since it includes a CarHMM, hierarchical HMM, and DFT-based transformation. Figure \ref{fig:CarHHMM-DFT} shows the corresponding graphical model. The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above. See the appendix for details of likelihood evaluation. In addition to the CarHHMM-DFT, we consider three similar variations for comparison. As in the full model, all components of $\Z_{t,t^*}$ are assumed to be independent when conditioned on dive type and subdive state for each of the following models:

@@

We use two visual tools to evaluate the CarHHMM-DFT: pseudoresidual plots and empirical histograms. The pseudoresidual of an observation $y_t$ for a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution. If the model is correct, then all pseudoresiduals are independent and follow a standard Normal distribution. Histograms of the pseudoresiduals mostly support that this model is well-specified. One exception is $\Ztwo_{t,t^*}$, whose pseudoresiduals are noticeably right-skewed (see Figure \ref{fig:pseudoresids}). This implies that the true distribution of $\Ztwo_{t,t^*}$ may follow a heavier-tailed distribution than the Gamma distribution used in the case study. See Sections 1.4 through 1.6 of the supplementary material for pseudoresidual plots for all observations.

We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:empirical_dist}. Each observed dive duration is weighted by the decoded probability that it corresponds to a particular dive type. This results in two histograms - one corresponding to dive type 1 and another corresponding to dive type 2. Each histograms is then plotted with the corresponding emission distribution learned by the CarHHMM-DFT. See Sections 1.4 through 1.6 of the supplementary material for analogous histograms corresponding to the fine-scale observations $\z_{t,t^*}$. Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions. In particular, $\Ztwo_{t,t^*}$ is right-skewed and $\Zone_{t,t^*}$ has heavy tails corresponding to subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.

>>>>

We then select a model corresponding to the fine-scale observations of acceleration. Similarly to the coarse model, we rely on lag plots and visual inspection to select $N^*=3$. Although $N^*$ is selected heuristically, we test the validity of this model using techniques in Section \ref{subsec:model_validation}.
%
In contrast to the coarse-scale observations, the fine-scale acceleration data exhibit significant sinusoidal behaviour. Thus, we transform each fine-scale observation sequence $y_t^*$ into $\z_t$ using Equation (\ref{eqn:z}) with window size $h=100$ (two seconds) and maximum frequency $\tilde{\omega}=10$ (5 Hz).
%We select a window size of two seconds since it appears be the smallest window size that captures at least one full period of each ``wiggle" in the acceleration data. 
We then have that $\z_{t,\tilde t^*} = \left\{\zone_{t,\tilde t^*},\ztwo_{t,t^*}\right\}$, where $\zone_{t,t^*}$ is a three dimensional vector of component-wise average acceleration and $\ztwo_{t,t^*}$ is a scalar describing the ``wiggliness" of window $\tilde t^*$. Even after transforming the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material). Therefore, we choose a CarHMM as defined in Section 2.2 as the fine-scale model.

@@

All components of $\Zone_{t,\tilde t^*}$ are assumed to be Normally distributed as in Equation (\ref{eqn:carhmm}) for all $t$ and $\tilde t^*$. Each component is assumed to have its own mean and variance parameters, but all components have the same auto-correlation parameter. 

@@

Given that $X^*_{t,t^*} = i^*$, $\Ztwo_{t,\tilde t^*}$ is assumed to follow a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$. We do not include $\Ztwo_{t,\tilde t^*-1}$ in the distribution of $\Ztwo_{t,\tilde t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state. 

@@

Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $\Z_{t,t^*}$ when $X^*_{t,t^*} = i^*$. 

@@

We refer to this final model as the \textbf{CarHHMM-DFT} since it includes a CarHMM, hierarchical HMM, and DFT-based transformation. Figure (\ref{fig:CarHHMM-DFT}) shows the corresponding graphical model. The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above. See the appendix for details of likelihood evaluation. In addition to the CarHHMM-DFT, we consider three similar variations for comparison. As in the full model, all components of $\Z_{t,t^*}$ are assumed to be independent when conditioned on dive type and subdive state for each of the following models:

@@

We use two visual tools to evaluate the CarHHMM-DFT: pseudoresidual plots and empirical histograms. The pseudoresidual of an observation $y_t$ for a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution. If the model is correct, then all pseudoresiduals are independent and follow a standard Normal distribution. Histograms of the pseudoresiduals mostly support that this model is well-specified. One exception is $\Ztwo_{t,t^*}$, whose pseudoresiduals are noticeably right-skewed (see Figure \ref{fig:pseudoresids}). This implies that the true distribution of $\Ztwo_{t,\tilde t^*}$ may follow a heavier-tailed distribution than the Gamma distribution used in the case study. See Sections 1.4 through 1.6 of the supplementary material for pseudoresidual plots for all observations.

We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:empirical_dist}. Each observed dive duration is weighted by the decoded probability that it corresponds to a particular dive type. This results in two histograms - one corresponding to dive type 1 and another corresponding to dive type 2. Each histograms is then plotted with the corresponding emission distribution learned by the CarHHMM-DFT. See Sections 1.4 through 1.6 of the supplementary material for analogous histograms corresponding to the fine-scale observations $\z_{t,\tilde t^*}$. Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions. In particular, $\Ztwo_{t,\tilde t^*}$ is right-skewed and $\Zone_{t,\tilde t^*}$ has heavy tails corresponding to subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.

### 2020-11-13 Report/Sections/simulation.tex
shorten intro

%\section{Simulation Study}
We perform a simulation study based on data generated from the full CarHHMM-DFT as defined in Section \ref{subsec:model_selection} to evaluate each candidate model when the ground-truth is known. The parameters used to generate the data are loosely based on those estimated in the case study (see Table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity. In particular, we set $\Gamma$ such that the simulated data have equal numbers of short and long dives in expectation, the number of subdive states is set to $N^*=2$, and $\Zone_{t,t^*}$ is set to a scalar rather than a three dimensional vector. We then fit the data with four models: the full CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT. Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, empirical standard errors of parameter estimates, and fitting times. To assess the accuracy of uncertainty estimates, we also compare the empirical standard errors of a given model's parameter estimates with the standard errors estimated from the inverse of the observed Fisher information under that model.

@@

We generate 500 independent data sets from the CarHHMM-DFT. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
%
Dive duration is Gamma distributed, and the coarse-scale emission parameters $\theta$ are $\mu^{(1)} = 20s$, $\sigma^{(1)} = 5s$, $\mu^{(2)} = 80s$, and $\sigma^{(2)} = 25s$. After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2T^*_t$ seconds of each dive are ignored. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left\{X^*_{t,1}, \ldots \right\}$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}$$ 
%
for dive types 1 and 2, respectively.
Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $Z^*_{t,t^*} = \left\{\Zone_{t,t^*}, \Ztwo_{t,t^*}\right\}$ as defined in Equation (\ref{eqn:z}). Recall that $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ are independent when conditioned on all dive types and subdive states. Each observation of acceleration $\Zone_{t,t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,t^*}$ is Gamma-distributed without explicit auto-correlation. In particular, we set the distributions of $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ with the following parameters:
%
\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.05s, \enspace \phi_A^{*(\cdot,1)} = 0.99, \\
    %
    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.1s, \enspace \phi_A^{*(\cdot,2)} = 0.95, \\
    %
    \mu_W^{*(\cdot,1)} = 10.10, \quad \sigma_W^{*(\cdot,1)} = 3.18, \\
    %
    \mu_W^{*(\cdot,2)} = 305.94, \quad \sigma_W^{*(\cdot,2)} = 17.49.
\end{gather*}
%
It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows one realization of $Z^*$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.

@@

We use several metrics to compare the candidate models, including its accuracy when decoding dive types and subdive states (see Table \ref{table:accuracy}). If $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define average dive decoding accuracy as the average value of $\Pr(X_{t} = x_{t}|Y,Y^*)$ across all simulated data sets and dives. Analogously, if $X^*_{t,t^*}$ is the unknown subdive state associated with a particular model and $x^*_{t,t^*}$ is the true simulated subdive state for dive $t$ and window $t^*$, then we define the average subdive decoding accuracy as $\Pr(X^*_{t,t^*} = x^*_{t,t^*}|Y,Y^*)$ averaged across all simulated data sets, dives, and windows. We find both the average dive decoding accuracy and the average subdive decoding accuracy using the forward-backward algorithm.

>>>>

%\section{Simulation Study}
We perform a simulation study based on data generated from the full CarHHMM-DFT as defined in Section \ref{subsec:model_selection} to evaluate each candidate model when the ground-truth is known. The parameters used to generate the data are loosely based on those estimated in the case study (see Table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity. In particular, we set $\Gamma$ such that the simulated data have equal numbers of short and long dives in expectation, the number of subdive states is set to $N^*=2$, and $\Zone_{t,\tilde t^*}$ is set to a scalar rather than a three dimensional vector. We then fit the data with four models: the full CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT. Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, empirical standard errors of parameter estimates, and fitting times. To assess the accuracy of uncertainty estimates, we also compare the empirical standard errors of a given model's parameter estimates with the standard errors estimated from the inverse of the observed Fisher information under that model.

@@

We generate 500 independent data sets from the CarHHMM-DFT. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
%
Dive duration is Gamma distributed, and the coarse-scale emission parameters $\theta$ are $\mu^{(1)} = 20s$, $\sigma^{(1)} = 5s$, $\mu^{(2)} = 80s$, and $\sigma^{(2)} = 25s$. After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each dive are ignored. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left\{X^*_{t,1}, \ldots \right\}$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}$$ 
%
for dive types 1 and 2, respectively.
Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$ as defined in Equation (\ref{eqn:z}). Recall that $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states. Each observation of acceleration $\Zone_{t,\tilde t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,\tilde t^*}$ is Gamma-distributed without explicit auto-correlation. In particular, we set the distributions of $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ with the following parameters:
%
\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.05s, \enspace \phi_A^{*(\cdot,1)} = 0.99, \\
    %
    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.1s, \enspace \phi_A^{*(\cdot,2)} = 0.95, \\
    %
    \mu_W^{*(\cdot,1)} = 10.10, \quad \sigma_W^{*(\cdot,1)} = 3.18, \\
    %
    \mu_W^{*(\cdot,2)} = 305.94, \quad \sigma_W^{*(\cdot,2)} = 17.49.
\end{gather*}
%
It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows one realization of $Z^*$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.

@@

We use several metrics to compare the candidate models, including its accuracy when decoding dive types and subdive states (see Table \ref{table:accuracy}). If $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define average dive decoding accuracy as the average value of $\Pr(X_{t} = x_{t}|Y,Y^*)$ across all simulated data sets and dives. Analogously, if $\tilde X^*_{t,\tilde t^*}$ is the unknown subdive state associated with a particular model and $\tilde x^*_{t,\tilde t^*}$ is the true simulated subdive state for dive $t$ and window $\tilde t^*$, then we define the average subdive decoding accuracy as $\Pr(\tilde X^*_{t,\tilde t^*} = \tilde x^*_{t,\tilde t^*}|Y,\tilde Y^*)$ averaged across all simulated data sets, dives, and windows. We find both the average dive decoding accuracy and the average subdive decoding accuracy using the forward-backward algorithm.
